Introducing an Ecological Trait-data Standard

### 2017-09-22 manuscript/manuscript.Rmd
Comments and edits on the Introduction

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->
<!--Nadja: I would not mention the distinction between effect and response traits-->
Traits have also proven useful for bridging individual level behaviour and physiology to processes at the ecosystem scale [@diaz13], for understanding physiological and evolutionary trade-offs, and for identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15]. Inferring an unknown trophic role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
<!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->

On the level of communities, community weighted means or variance of functional traits are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].

Traits furthermore allow comparison across trophic levels. Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.   
<!--Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->

A global synthesis of traits can bring major insights in ecosystem processes and enable projections in future climate and land use scenarios. Traits link ecological research to evolution, to management (functions and services) for biodiversity conservation (land use intensification), and to climate research (environmental filters). 
<!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
It is thus an interdisciplinary/cross-topic field with great potential.

@@

However, an integration of several studies across trophic levels, taxa or organisational scales remains difficult because of different organismal, methodological, and functional focus. A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and make data use possible for future generations.

@@

Summarize different questions, currently proposed in above- and below ground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.

@@

Trait data have already been standardised into centralised databases for specific organism groups and research questions. Most prominently, plant traits from many different sources --including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08]-- have been collated into the quickly growing TRY database [@kattge11]. TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] or root measurements [Fine-root Ecology Database; @iversen17]. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism groups or other trait-based research questions would be desireable.

Microbial ecology also makes great use of trait-based approaches to asses genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->

In the animal kingdom, specialised trait-databases have been created that cover different organism groups or interaction types or ecosystems. For instance, The BETSI database collects traits on soil invertebrats (http://betsi.cesab.org/).

@@

Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g. @Gossner2015SData,@Gossner2015Ecology,@Renner2014]. It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus. Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts produced for plant traits (@refs).

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand. Additionally, the training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization.

@@

Such ontologies are being developed for plant traits, for instance in the TOP Thesaurus of plant traits that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].

@@

As a minimum consensus, trait datasets may be defined as follows: 
A trait-dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an individual or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species or higher taxon). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon. 
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'.

@@

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism. We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species. This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.

Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns. Computationally more effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID. <!-- Malte: I think the difference between the long-table format and the previous one is not clear enough. Nadja: Is this better now? -->
The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.

@@

>>>>

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07] and have been used to describe the role of species in ecosystems or their ability to persist under certain environmental conditions (@refs).
<!--^[Some frameworks differentiate effect traits and response traits (or performance and fitness traits), depending on the research question [refs, @violle07]]-->
<!--Nadja: I would not mention the distinction between effect and response traits-->
Traits have also proven useful for bridging individual level behaviour and physiology to processes at the ecosystem scale [@diaz13], for understanding physiological and evolutionary trade-offs, and for identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15]. Inferring an unknown trophic role or ecological function of individuals from their apparent features [@duarte11; @fontaine05] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and have not been observed in the ecosystem.
<!--Nadja: I am not sure if I understand what you mean with "have not been observed in the ecosystem".-->
<!--Malte: I guess, you mean their behaviour has not been observed, i.e. might know what they look like (what traits they have) because we caught / found individuals, but we have no idea what they do, where / how they live, with whom they interact etc. How about "... are yet undescribed and their behaviour, their interactions with other organisms and the environment and their ecological function are virtually unknown."-->

On the level of communities, community weighted means or variance of functional traits are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @hopfenmuller14; @jennings09; @moretti13; @bello10; @vandewalle10; @Gossner2016; @Simons2015Ecology; @Gamez-Virues2015; @Birkhofer2014]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem (@refs) or of ecosystem services [@wood15; @bello10].
<!--Malte: It is not only CWM's and variance of traits, but also functional diversity metrics that are commonly used to understand change sin functional composition. I would probably add that here.-->

Traits furthermore allow comparison across trophic levels. Just as for pollinator and plant trait-pairs, resource defense traits and predator feeding traits have co-evolved. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens which may be used to inform ecological networks [@munoz17; @schleuning] and parameterize ecological population-dynamic models [@refs].This is also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups.   
<!--Even at the macroecological scale, adaptions against large herbivore grazing pressure have been shown in global patterns of plant defense traits [@diaz07]. 
Nadja: I am not sure how this fits here as it is very specific. Maybe phrase it more generally?-->
<!--Malte: First sentence, 'comparison across trophic levels". Not sure if this is the ideal wording; is it really comparisons that they allow for here? What about "Functional traits furthermore are an important tool for studying interactions across trophic levels." -->

Traits link ecological research to evolution, management (functions and services) for biodiversity conservation (land use intensification), and climate research (environmental filters). As such, a global synthesis of traits would offer major insights in ecosystem processes and enable projections in future climate and land use scenarios. 
<!--Nadja: Not sure if my changes make sense. I also think that the brackets make this sentence too complicated.-->
Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential.

@@

Despite the obvious benefit of such global trait synthesis, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus. A harmonization of trait data standards would be of huge benefit for the advancement of trait-based research and enable efficient data use for future generations.

@@

Summarize different questions, currently proposed in above- and belowground work within the Exploratories, focus on plant traits, arthropod traits, microbial traits.

@@

Trait data have already been standardised into centralised databases for specific organism groups and research questions. Most prominently, plant traits from many different sources --including the extensive LEDA traitbase for the Northwest-European flora [@kleyer08]-- have been collated into the quickly growing TRY database [@kattge11]. TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] or root measurements [Fine-root Ecology Database; @iversen17]. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism groups or other trait-based research questions would be desirable.

Microbial ecology also makes great use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14]. <!-- more on microbial databases -->

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, The BETSI database collects traits on soil invertebrates (http://betsi.cesab.org/).

@@

Besides these databases, an unknown plethora of single datasets has been published along with article publications over the past decades [e.g., @Gossner2015SData,@Gossner2015Ecology,@Renner2014]. It becomes apparent, that different databases and individual datasets come with their own structure, reflecting the research questions of the initiatives behind them and their organismal focus. Additionally, it has been pointed out that data on animals --covering multiple trophic levels and having evolved very different body plans-- will not be a simple task and go far beyond the efforts made for plant trait databases (@refs).

Public research funding agencies increasingly demand the publication of data without access restrictions [@swan12; @allison15], but not all scientific databases for functional traits are compliant with this demand. Additionally, the training on compliance with open data standards within the community of environmental sciences can still be much improved [@schmidt16]. At the same time, the open-access publications in low-threshold data repositories such as figshare, datadryad, researchgate or zenodo are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

@@

Such ontologies are being developed for plant traits, for instance in the TOP, Thesaurus of plant traits, that is used for the TRY database [@garnier17] and the Planteome database and ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches or the 'Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits' [@moretti17].

@@

As a minimum consensus, trait datasets may be defined as follows: 
A trait-dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an individual or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon. 
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'.

@@

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation and sexual dimorphism. We term these tables as occurrence data tables, since each row is centered around a single physical occurrence of a species.<!-- Malte: ... Or a single measurement, right? --> This format is the most intuitive for recording own empirical measurements and therefore is common for measured data and rarely found for reported facts obtained from the literature.

Both the matrix format and the occurrence data format have the different trait values which have been collected or measured on one entity stored in multiple columns. Computationally more effective and allowing for highest flexibility is the storage of traitdata in long-table formats, where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) via a unique identifier which is assigned to a taxon. This allows for repeated measurements, even on a single individual. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID. <!-- Malte: I think the difference between the long-table format and the previous one is not clear enough. Nadja: Is this better now? Malte: Yes, it is. -->
The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID.

@@

M.J. was supported by the German Research Foundation within the framework of the Jena Experiment (FOR 1451) and by the Swiss National Science Foundation.
